Analysis of competitive surfing tournaments with generalized Bradley-Terry

### 2021-08-25 inst/surfing_jsa.tex
start work on JSA house style

\documentclass{article}
\usepackage{amssymb}
\usepackage{graphicx}
\begin{document}
\bibliographystyle{plain}

@@

\section*{Abstract}
Here we analyse competitive surfing, specifically the 2018 World Surf
League, using formal statistical methods.  We use generalized
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Specifically, we quantify the dominance of
the top competitors using likelihood techniques, and go on to study
the ``Brazilian storm'' phenomenon using reified entities in two ways.
Firstly we assess the supposed Brazilian preference for point and
beach break beaches; and secondly we consider tournament results from
the perspective of tournament theory and test for competitors
modifying their strategy in the presence of compatriot rivals.  We
quantify the evidence for these commonly assumed features of
contemporary competitive surfing and suggest further avenues of
research.

@@

Surfing is a popular and growing activity in terms of both
participation and viewing \cite{warshaw2010}.  Competitive surfing
involves surfers competing against one or two other surfers in heats
lasting 30-50 minutes \cite{booth1995}.  Points are awarded on a
10-point scale: each of five judges awards points to a surfer riding a
particular wave.  The World Surf League (WSL) is the main governing
body of surfing competitions \cite{wsl}.  The WSL conducts a world tour in
which the best ranked 34 surfers compete; in addition, at each tour
venue, two "wildcard" surfers also enter the competition, who are
ignored here.  The 2019 WSL tour saw competitive events held at eleven
different surf locations.

@@

Farley {\em et. al.} \cite{farley2013} analyse surfing results using
one-way ANOVA techniques and report comparisons between the top 10 and
bottom 10 surfers in the 2013 World Championship Tour and conclude
that the top-ranked athletes are more consistent than the lower-ranked
group.  They report that ``to date, only a limited number of research
studies have reported on the results of competitive surfing" [p44],
and the analysis presented here fills some of this gap.

Other statistical research into competitive surfing appears to focus
on the scoring of aerial manoeuvres compared with other manoeuvres in
events.  Lundgren {\em et. al.} \cite{lundgren2014}, for example,
report that aerial manoeuvres scored higher than other stunts but had
a lower completion rate, a theme we return to below.

@@

The Bradley-Terry model \cite{bradley1954} assigns non-negative strengths
$p_1,\ldots, p_n$ to each of $n$ competitors in such a way that the
probability of $i$ beating $j\neq i$ in pairwise competition is
$\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that $\sum
p_i=1$.  Further, we use a generalization due to \cite{luce1959}, in which
the probability of competitor~$i$ winning in a field of $\left\lbrace
1,\ldots, n\right\rbrace$ is $\frac{p_i}{p_1+\cdots +p_n}$.  Noting
that there is information in the whole of the finishing order, and not
just the first across the line, we can follow \cite{plackett1975} and
consider the runner-up to be the winner among the remaining
competitors, and so on down the finishing order. Without loss of
generality, if the order of finishing were $1,2,3,4,5$, then a
suitable Plackett-Luce likelihood function would be

@@

and this would be a forward ranking Plackett-Luce model in the
terminology of \cite{johnson2020}.  We now use a technique due to Hankin
(2010, 2017) and introduce fictional (reified) entities whose nonzero
Bradley-Terry strength helps certain competitors or sets of
competitors under certain conditions.  The original example was the
home-ground advantage in football.  If players (teams) $1,2$ with
strengths $p_1,p_2$ compete, and if our observation were $a$ home wins
and $b$ away wins for team $1$, and $c$ home wins and $d$ away wins
for team~$2$, then a suitable likelihood function would be

@@

\noindent where $p_H$ is a quantification of the beneficial home
ground effect.  Similar techniques have been used to account for the
first-move advantage in chess, and here we apply it to assess the
supposed Brazilian preference for reef-breaking waves.

@@

Paired comparisons are particularly favourable to Bradley-Terry
because the likelihood function is exponential family and as such is
amenable to analysis via general linear models~\cite{turner2012}.
However, this is not applicable when more than two entities are
compared; and indeed here we see comparisons between three surfers
and, in the case of Jeep Surf Freshwater Pro (wct08), many more.  It
is natural to ask how much information is contained in the
observations where more than two surfers are in the water.

@@

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
  hyper2} software~\cite{hankin2017} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  The Method of Support \cite{edwards1992}
rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
There is thus strong evidence to suggest that the competitors are
indeed of different skills and the result actually reflects genuine
differences in abilities, rather than the result of random chance.
Alternatively, we may view the dataset as being sufficiently
informative to reveal actual differences in surfing ability.  Observe
that the competitors' points (that is, the final column of
table~\ref{resultstable}) cannot be used for this type of
probabalistic analysis: the points awarded are arbitrary and changing
the points system would change the points scored but not affect the
competitors' behaviour.

@@

We may ask whether the top two officially ranked competitors, Medina
and Ferreira, have different strengths.  To test the null of equal
strength, we maximize the likelihood subject to the null, and compare
with the global maximum likelihood~\cite{hankin2010}.  This gives us a
support difference of about 0.05, well short of Edwards's 2 units of
support criterion.  Alternatively, we could observe that the
$-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the
tail region of its asymptotic $\chi^2_1$ distribution.  We may thus
assert that there is no strong evidence for Ferreira's actually having
a competitive advantage over Medina.  We may similarly wonder whether
Medina is equal in strength to Wilson; here we reject the null with
support difference of about 2.33, corresponding to a $p$-value of
$0.031$.

@@

In surfing lore, the ``Brazilian storm'' refers to the emergence of
Brazil as a dominant nation of elite surfers.  The term was first used
to describe the fact that six out of eight querterfinalists at the 2011 Nike
Lowers Pro were Brazilian; it rose in public conciousness following
the victories of Medina and de Souza in the World Surf League
championships in 2014 and 2015 respectively.  The storm shows no signs
of abating, with commentators such as Douglas opining that this
generation ``represents a seismic shift for Brazilian competitors onto
an entirely new plane'' \cite{douglas2020}.

@@

The statistical method described here can test many different
hypotheses inspired by surfing lore.  For example, one frequent
suggestion~\cite{burgess2020,ho2021} is that Brazilian surfers tend to
have skillsets that favour point and beach break wave types, readily
accessible to these surfers' native Brazil; see~\cite{scarfe2003} for
a scientific overview of surf break characteristics,
and~\cite{butt2004} for a surfer's perspective.

@@

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics~\cite{tulloch2019}, and it is natural to ask
whether this would affect the competition in some way~\cite{ho2021}.
One suggestion would be that competitors become more focussed on
beating compatriots than surfers from other countries, the thinking
being that admission to a country's Olympic team would be of more
importance to a competitor than their overall standing in the
tournament.  One might expect competitors to value their ranking when
restricted to within their country more than their international
ranking.  If this is the case then we would expect competitors to
change their behaviour if they are in the water with another surfer of
the same nationality.  Specifically, if we have three surfers in the
water of whom two are of one nationality and the third of another, we
would expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.  Such hypotheses are difficult to test with conventional
techniques as competitors's strengths themselves are nuisance
parameters~\cite{basu1977}.

@@

\bibliography{chess}
\end{document}

>>>>

\documentclass{article}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{natbib}

  \author{Driver, Timothy${}^{1,2}$ \and
    Hankin, Robin K. S.${}^{2,3}$ \\
    \texttt{hankin.robin@gmail.com}\\(corresponding author)}

  \date{${}^1$ {\bf Tim what do you want here you can put anything}\\
    ${}^{1,2}$ Auckland University of Technology
    Wellesley Street, Auckland, New Zealand\\
    +64-9-9219999x5106\\
    ${}^3$ Corresponding author}

@@

\begin{document}

  \maketitle

\subsubsection*{Abstract}
   Here we analyse competitive surfing, specifically the 2018 World Surf
League, using formal statistical methods.  We use generalized
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Specifically, we quantify the dominance of
the top competitors using likelihood techniques, and go on to study
the ``Brazilian storm'' phenomenon using reified entities in two ways.
Firstly we assess the supposed Brazilian preference for point and
beach break beaches; and secondly we consider tournament results from
the perspective of tournament theory and test for competitors
modifying their strategy in the presence of compatriot rivals.  We
quantify the evidence for these commonly assumed features of
contemporary competitive surfing and suggest further avenues of
research.
\newpage

@@

Surfing is a popular and growing activity in terms of both
participation and viewing \citep{warshaw2010}.  Competitive surfing
involves surfers competing against one or two other surfers in heats
lasting 30-50 minutes \citep{booth1995}.  Points are awarded on a
10-point scale: each of five judges awards points to a surfer riding a
particular wave.  The World Surf League (WSL) is the main governing
body of surfing competitions \citep{wsl}.  The WSL conducts a world tour in
which the best ranked 34 surfers compete; in addition, at each tour
venue, two ``wildcard" surfers also enter the competition, who are
ignored here.  The 2019 WSL tour saw competitive events held at eleven
different surf locations.

@@

\citet{farley2013} analyse surfing results using
one-way ANOVA techniques and report comparisons between the top 10 and
bottom 10 surfers in the 2013 World Championship Tour and conclude
that the top-ranked athletes are more consistent than the lower-ranked
group.  They report that ``to date, only a limited number of research
studies have reported on the results of competitive surfing" [p44],
and the analysis presented here fills some of this gap.

Other statistical research into competitive surfing appears to focus
on the scoring of aerial manoeuvres compared with other manoeuvres in
events.  \citet{lundgren2014}, for example,
report that aerial manoeuvres scored higher than other stunts but had
a lower completion rate, a theme we return to below.

@@

The Bradley-Terry model \citep{bradley1954} assigns non-negative strengths
$p_1,\ldots, p_n$ to each of $n$ competitors in such a way that the
probability of $i$ beating $j\neq i$ in pairwise competition is
$\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that $\sum
p_i=1$.  Further, we use a generalization due to \citep{luce1959}, in which
the probability of competitor~$i$ winning in a field of $\left\lbrace
1,\ldots, n\right\rbrace$ is $\frac{p_i}{p_1+\cdots +p_n}$.  Noting
that there is information in the whole of the finishing order, and not
just the first across the line, we can follow \citet{plackett1975} and
consider the runner-up to be the winner among the remaining
competitors, and so on down the finishing order. Without loss of
generality, if the order of finishing were $1,2,3,4,5$, then a
suitable Plackett-Luce likelihood function would be

@@

and this would be a forward ranking Plackett-Luce model in the
terminology of \citet{johnson2020}.  We now use a technique due to
\citet{hankin2010,hankin2017} and introduce fictional (reified)
entities whose nonzero Bradley-Terry strength helps certain
competitors or sets of competitors under certain conditions.  The
original example was the home-ground advantage in football.  If
players (teams) $1,2$ with strengths $p_1,p_2$ compete, and if our
observation were $a$ home wins and $b$ away wins for team $1$, and $c$
home wins and $d$ away wins for team~$2$, then a suitable likelihood
function would be

@@

\noindent where $p_H$ is a quantification of the beneficial home
ground effect.  Similar techniques have been used to account for the
first-move advantage in chess~\citep{hankin2020}, and here we apply it
to assess the supposed Brazilian preference for reef-breaking waves.

@@

Paired comparisons are particularly favourable to Bradley-Terry
because the likelihood function is exponential family and as such is
amenable to analysis via general linear models~\citep{turner2012}.
However, this is not applicable when more than two entities are
compared; and indeed here we see comparisons between three surfers
and, in the case of Jeep Surf Freshwater Pro (wct08), many more.  It
is natural to ask how much information is contained in the
observations where more than two surfers are in the water.

@@

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
  hyper2} software~\citep{hankin2017} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  The Method of Support \citep{edwards1992}
rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
There is thus strong evidence to suggest that the competitors are
indeed of different skills and the result actually reflects genuine
differences in abilities, rather than the result of random chance.
Alternatively, we may view the dataset as being sufficiently
informative to reveal actual differences in surfing ability.  Observe
that the competitors' points (that is, the final column of
table~\ref{resultstable}) cannot be used for this type of
probabalistic analysis: the points awarded are arbitrary and changing
the points system would change the points scored but not affect the
competitors' behaviour.

@@

We may ask whether the top two officially ranked competitors, Medina
and Ferreira, have different strengths.  To test the null of equal
strength, we maximize the likelihood subject to the null, and compare
with the global maximum likelihood~\citep{hankin2010}.  This gives us a
support difference of about 0.05, well short of Edwards's 2 units of
support criterion.  Alternatively, we could observe that the
$-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the
tail region of its asymptotic $\chi^2_1$ distribution.  We may thus
assert that there is no strong evidence for Ferreira's actually having
a competitive advantage over Medina.  We may similarly wonder whether
Medina is equal in strength to Wilson; here we reject the null with
support difference of about 2.33, corresponding to a $p$-value of
$0.031$.

@@

In surfing lore, the ``Brazilian storm'' refers to the emergence of
Brazil as a dominant nation of elite surfers.  The term was first used
to describe the fact that six out of eight querterfinalists at the 2011 Nike
Lowers Pro were Brazilian; it rose in public conciousness following
the victories of Medina and de Souza in the World Surf League
championships in 2014 and 2015 respectively.  The storm shows no signs
of abating, with commentators such as Douglas opining that this
generation ``represents a seismic shift for Brazilian competitors onto
an entirely new plane'' \citep{douglas2020}.

@@

The statistical method described here can test many different
hypotheses inspired by surfing lore.  For example, one frequent
suggestion~\citep{burgess2020,ho2021} is that Brazilian surfers tend to
have skillsets that favour point and beach break wave types, readily
accessible to these surfers' native Brazil; see~\citet{scarfe2003} for
a scientific overview of surf break characteristics,
and~\citet{butt2004} for a surfer's perspective.

@@

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics~\citep{tulloch2019}, and it is natural to ask
whether this would affect the competition in some way~\citep{ho2021}.
One suggestion would be that competitors become more focussed on
beating compatriots than surfers from other countries, the thinking
being that admission to a country's Olympic team would be of more
importance to a competitor than their overall standing in the
tournament.  One might expect competitors to value their ranking when
restricted to within their country more than their international
ranking.  If this is the case then we would expect competitors to
change their behaviour if they are in the water with another surfer of
the same nationality.  Specifically, if we have three surfers in the
water of whom two are of one nationality and the third of another, we
would expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.  Such hypotheses are difficult to test with conventional
techniques as competitors's strengths themselves are nuisance
parameters~\citep{basu1977}.

@@

\bibliographystyle{agsm}
\bibliography{chess}
\end{document}
